---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


####Load the library's used
I will be suppressing the messages generated during load. I have reviewed these messages and they provide no value. You may want to remove the message suppression when running on your system.
```{r}
suppressWarnings(suppressMessages(library(dplyr)))
suppressWarnings(suppressMessages(library(ggplot2)))
```

## Loading and preprocessing the data
```{r}

activity <- read.table(unz("activity.zip","activity.csv"),header = T,quote="\"", sep=",",na.strings = NA)
```

```{r}

activity$date <- as.Date(activity$date, format = "%Y-%m-%d ")
activity$interval <- as.factor(activity$interval)
activity$steps <- as.numeric(activity$steps)

activity <- cbind(activity,intervalgroup = as.integer(levels(activity$interval)) %/% 12)

```



## What is mean total number of steps taken per day?
```{r}
activity_stepday <- aggregate(steps ~ date, activity, sum)
colnames(activity_stepday) <- c("date","steps")

mean_steps <- format(mean(activity_stepday$steps, na.rm=TRUE),digits=2, nsmall=2, big.mark=",",small.mark=".")
median_steps <- format(median(activity_stepday$steps, na.rm=TRUE),digits=2, nsmall=2, big.mark=",",small.mark=".")

```
#####The mean number of steps per day is `r mean_steps`
#####The median number of steps per day is `r median_steps`

######If we were to plot it out, it would looks like this
```{r}
ggplot(activity_stepday, aes(x = date, y=steps)) + 
         geom_histogram(color="blue",stat="identity") +
        labs(title="steps Taken per Day", 
             x = "Day", y = "Steps") 
```

## What is the average daily activity pattern?
```{r}
mean_stepsmean_intervalsteps <- aggregate(activity$steps,by=list(interval=activity$interval), 
                       FUN=mean, 
                       na.rm=TRUE)
                     
                       
names(mean_stepsmean_intervalsteps)<-c("interval","steps")
mean_stepsmean_intervalsteps$interval <- as.integer(levels(mean_stepsmean_intervalsteps$interval)[mean_stepsmean_intervalsteps$interval])
max_steps = max(mean_stepsmean_intervalsteps$steps)
max_interval= filter(mean_stepsmean_intervalsteps, steps==max_steps)$interval

ggplot(mean_stepsmean_intervalsteps, aes(x=interval, y=steps)) +   
        geom_line(color="blue") +  
        labs(title="Average Daily Step Pattern", x="interval", y="Number of steps")

```

####The interval that had the maximum average steps is `r max_interval` with a daily average of `r format(max_steps,digits=2, nsmall=2, big.mark=",",small.mark=".")` steps.

## Imputing missing values

1.) The total number of missing rows can be easily determined by the getting the number of rows where the is.na is true in the steps column
```{r}
missing_rows <-format(nrow(activity[is.na(activity$steps),]),digits=2, nsmall=2, big.mark=",",small.mark=".")
```

#####There are `r missing_rows` rows with NA values in the steps column

2.) To patch up the NAs is straight forward using the dplyr package. 

-group by interval 

-create a Mean col by interval 

-replace the NA's with the Mean 

-loose the temp column Mean 

This will be done in one piped statement

```{r}
activityfilled <- activity %>%
       group_by(interval) %>%
       mutate(Mean= mean(steps, na.rm=TRUE), 
               steps = replace(steps, 
                  which(is.na(steps)), first(Mean))) %>%
       select(-Mean)
```


```{r}
missing_rows <-format(nrow(activityfilled[is.na(activityfilled$steps),]),digits=2, nsmall=2, big.mark=",",small.mark=".")
```

#####After imputing the data, the resulting set shows `r missing_rows` rows with NA values in the steps column


Here I will create an average steps per day sum on the filled data set, and plot it.

```{r}
activityfilled_stepday <- aggregate(steps ~ date, activityfilled, sum)
colnames(activityfilled_stepday) <- c("date","steps")
ggplot(activityfilled_stepday, aes(x = date, y=steps)) + 
         geom_histogram(color="blue",stat="identity") +
        labs(title="steps Taken per Day", 
             x = "Day", y = "Steps") 
```


####Now that we have datasets to compare, let's look closer.


We calculate the mean and median of each. 

```{r}
activity_mean <- format(mean(activity_stepday$steps,na.rm = TRUE),digits=2, nsmall=2, big.mark=",",small.mark=".")
activity_median <- format(median(activity_stepday$steps,na.rm = TRUE),digits=2, nsmall=2, big.mark=",",small.mark=".")
activityfilled_mean <- format(mean(activityfilled_stepday$steps,na.rm = TRUE),digits=2, nsmall=2, big.mark=",",small.mark=".")
activityfilled_median <- format(median(activityfilled_stepday$steps,na.rm = TRUE),digits=2, nsmall=2, big.mark=",",small.mark=".")
```

The step mean of the original data set is `r activity_mean`, and the step mean of the filled data set is `r activityfilled_mean`.
The step median of the original data set is `r activity_median`, and the step median of the filled data set is `r activityfilled_median`.

## Are there differences in activity patterns between weekdays and weekends?
For my evaluation, I will use the intervalgroup column created earlier (grouping of 20 Intervals). I will graph the steps taken during the day for both weekdays and weekends. I will also compare the average steps per day for weekdays vs weekends. 

```{r}
 activity <- cbind(activity,weekday = tolower(weekdays(activity$date)))
 activity <- cbind(activity,daytype=ifelse((activity$weekday == "saturday" | 
                                     activity$weekday == "sunday"), "weekend", 
                                     "weekday"))
 we_stepday <- activity %>%
                 filter(daytype=="weekend")
 we_avg <- format(mean(aggregate(we_stepday$steps,by=list(date=we_stepday$date), 
                       sum, 
                       na.rm=TRUE)$x,na.rm=T),digits=2, nsmall=2)
 wd_stepday <- activity %>%
                 filter(daytype=="weekday")
 wd_avg <- format(mean(aggregate(wd_stepday$steps,by=list(date=wd_stepday$date), 
                       sum, 
                       na.rm=TRUE)$x,na.rm=T),digits=2, nsmall=2)

 activity_stepday <- aggregate(steps~interval+daytype, activity, mean)
 activity_stepday$interval <- as.integer(levels(activity_stepday$interval))
 ggplot(activity_stepday, aes(x=interval, y=steps)) + 
        geom_line(color="blue",) +
        facet_wrap(~ daytype, nrow=2, ncol=1) +
        labs(x="Interval", y="Number of steps") 

 
```       

#####Yes. It appears that there is a there is fewer steps per day on the weekday. On the weekday the average steps taken per day is `r wd_avg`. However on the weekend the average steps taken per day is `r we_avg`. Also activity appears to occur later in the day and more continuous on weekends       
       